---
title: "Question 1: Coffee Sourcing Recommendations"
author: "22761926"
date: "`r Sys.Date()`"
output: powerpoint_presentation
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(tidyverse)

list.files('code/', full.names = T, recursive = T) %>% 
    as.list() %>% walk(~source(.))

Coffee <- Load_Coffee("../data/Coffee/Coffee.csv")

Coffee <- Coffee %>% Extract_Region()

```

## Dataset Overview: 2095 Coffees, 422 Roasters, 19 Countries

```{r}

Dataset_Overview(Coffee) %>% knitr::kable(digits = 2)

```

## Best Countries by Rating

```{r, fig.width=8, fig.height=5.5}

Region_Rating_Plot(Coffee)

```

## Rating by Roast Strength

```{r, fig.width=7, fig.height=5.5}

Roast_Plot(Coffee)

```

## Best Value Coffees: Guatemala Leads

```{r}
Top_Value_Coffees(Coffee, Top_N = 10) %>% 
  mutate(Value_Score = round(Value_Score, 1)) %>% 
  head(5) %>% 
  knitr::kable(digits = 2, 
              col.names = c("Coffee", "Roaster", "Region", "Roast", 
                           "Rating", "Cost/100g ($)", "Value Score"))
```

## Best Value Coffees: Guatemala Leads (Continued)

```{r}
Top_Value_Coffees(Coffee, Top_N = 10) %>% 
  mutate(Value_Score = round(Value_Score, 1)) %>% 
  tail(5) %>% 
  knitr::kable(digits = 2, 
              col.names = c("Coffee", "Roaster", "Region", "Roast", 
                           "Rating", "Cost/100g ($)", "Value Score"))
```

## Flavour Descriptor Impact

```{r, fig.width=8, fig.height=5}

Keyword_Highlight_Plot(Coffee)

```

## Recommended Suppliers: Kakalove Cafe Combines Reliability and Value

```{r}
Top_Roasters(Coffee, Min_Coffees = 3, Top_N = 10) %>% 
  knitr::kable(digits = 1, 
              col.names = c("Roaster", "Avg Rating", "Avg Cost/100g ($)", "N Coffees"))
```

## Recommendations

- **Country**: Ethiopia is the strongest, best-supported choice (high rating, 313 reviews). Kenya and Colombia are also reliable alternatives.

- **Roast**: Light and Medium-Light roasts rate highest; ratings decline as roast darkens.

- **Suppliers**: Kakalove Cafe offers the best mix of reliability, rating and affordability. Hula Daddy Kona Coffee is a strong premium option if budget allows.

## Recommendations (Continued)

- **Value**: Guatemalan coffees repeatedly appear among the best value picks - useful for margin-conscious stocking.

- **Flavour profile**: "Balanced" in reviews is a positive signal; "velvety" is a mild cautionary flag.