---
title: "Netflix: Content Trends by Country"
author: "22761926"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(tidyverse)

options(dplyr.summarise.inform = F)

list.files('code/', full.names = T, recursive = T) %>% 
    as.list() %>% walk(~source(.))

Titles <- read_rds("../data/netflix/titles.rds")
Credits <- read_rds("../data/netflix/credits.rds")
Movie_Info <- read_csv("../data/netflix/netflix_movies.csv")
HBO_titles <- read_rds("../data/US_Baby_names/HBO_titles.rds")

CountryList <- c("United States", "India", "United Kingdom", "Spain", "Nigeria", "Japan")
df_movies <- Prepare_Movie_Data(Movie_Info, CountryList = CountryList)
df_movies_scored <- Join_Scores(df_movies, Titles)

Theme_Words <- Movie_Theme_Words()
Theme_Results <- Theme_Words %>% map_df(~Theme_Extractor(df_movies, ThemeWord = .))

Netflix_Genres <- Clean_Netflix_Genre(Movie_Info) %>% Fix_Netflix_Genre_Names()
HBO_Genres <- Clean_HBO_Genre(HBO_titles) %>% Align_HBO_Documentary()

```

# Overview

- We compared Netflix movie content across six countries: United States, India, United Kingdom, Spain, Nigeria and Japan.
- We looked at movie length, content maturity ratings, common themes in movie descriptions, and audience scores.
- As a bonus, we compared genre popularity between Netflix and HBO, reusing the HBO dataset from Question 2.

# Movie Length by Country

```{r, fig.width=7, fig.height=5}

Plot_Country_Duration(df_movies)

```

- India's films average 127 minutes, notably longer than every other country in this sample.
- This fits with the well-known convention of longer Bollywood runtimes, which often include multiple storylines and musical sequences.
- The United Kingdom has the shortest average runtime at 85 minutes, followed closely by the United States.
- Most countries cluster fairly close together, between 85 and 110 minutes, with India as the clear outlier.

# Content Maturity by Country

```{r, fig.width=8, fig.height=5.5}

Plot_Country_Maturity(df_movies)

```

- Spain's content skews heavily toward Mature ratings, at nearly 86%, the highest of any country in this sample.
- India sits at the opposite end. Mature content makes up only 24% of its movies, with Parental Guidance/Teen content dominating at 74%.
- The United States and United Kingdom sit roughly in the middle, both with Mature content making up around half of their movies.
- General/Family content is a small share everywhere, never exceeding 8% in any country.

# Movie Themes by Country

```{r, fig.width=9, fig.height=6}

Plot_Theme_Comparison(Theme_Results, CountryList = CountryList)

```

- "Love" and "friend" are the most commonly mentioned themes overall, particularly in India and Spain.
- India has the highest mentions of both "love" and "friend" across all six countries.
- The United States stands out for "comedy" mentions, the highest of the group.
- Darker themes such as murder, crime and revenge appear consistently less often than relationship-driven themes like love, friend and family, across every country in this sample.

# Movie Ratings by Country

```{r, fig.width=7, fig.height=5.5}

Plot_Country_Scores(df_movies_scored)

```

- IMDb and TMDB scores broadly agree with each other across countries, giving confidence that the pattern is real rather than noise from a single rating source.
- The United Kingdom and Japan show the highest average ratings, though both rely on small samples of 47 and 18 movies respectively, so these averages should be treated with some caution.
- India and the United States, our two largest and most reliable samples at 382 and 699 movies, sit close together at essentially the same average score, around 6.27 to 6.28.
- Nigeria has the lowest average rating of the six countries, at 5.77.

# Bonus: Comparing Netflix and HBO Genre Mix

```{r, fig.width=7, fig.height=5}

Plot_Platform_Genre(Netflix_Genres, HBO_Genres, Movie_Info, HBO_titles)

```

- HBO's catalogue is more concentrated than Netflix's across all three genres compared here.
- Drama is the largest category on both platforms, making up 43.8% of HBO's catalogue and 39.6% of Netflix's.
- Comedy and documentary follow the same pattern, with HBO's share exceeding Netflix's in both cases.
- This comparison was limited to drama, comedy and documentary, the genres with clearly comparable naming once pluralization and terminology differences between the two platforms were resolved. A full genre-by-genre comparison would require reconciling many more inconsistently named categories.